Intelligence and Age at First Intercourse: Cause or Confound?

### 2016-01-12 Common/content/writing/results_between.tex
incorporated joe's changes
more joe changes
fixed tables

\subsubsection{Gen1 Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Gen1 sister averages (NLSY79 mothers) of standardized AFQT scores were used to predict Gen2 averages of gender standardized AFI. Table \ref{table_Mean_Mom_Intelligence_Mean_Child_AFI_9} displays the results by Gen2 categories. The Mixed model reports the averages of the first born child (both males and females) of each sister (n $= 342$), the Daughters model reports the averages of the first born girls (n $= 264$), and the Sons model reports the averages of the first born sons (n $= 282$). All three models reveal similar results. A one unit increase in the average standardized intelligence of the children's mothers predicted a statistically significant increase of $\approx .013$ standard deviations in average Gen2 AFI. The adjusted R$^{2}$ varied slightly by Gen2 categories (Mixed $= .087$, Daughters $= .097$, Sons $= .103$).

\subsubsection{Gen2 Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Gen2 cousin averages of standardized intellectual ability scores were used to predict Gen2 averages of gender--standardized AFI. Table \ref{table_Mean_Child_Intelligence_Mean_Child_AFI_9} displays the results by Gen2 categories. The Mixed model reports the averages of the first-borns of each of the NLSY79 mothers (sisters) (n $= 344$), the Daughters model reports the averages of the first born girls (n $= 267$), and the Sons model reports the averages of the first born sons (n $= 283$). All three models reveal similar results. A one unit increase in the average standardized intelligence of the children predicted a statistically significant $\approx .075$ standard deviation increase in average Gen2 AFI. The adjusted R$^{2}$ varied slightly by Gen2 categories (Mixed $= .014$, Daughters $= .016$, Sons $= .009$). \cil{double checked. These R$^{2}$are correct.}

\subsubsection{Joint Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Results from the Gen1 sister averages of standardized AFQT scores and Gen2 cousin averages of standardized intellectual ability scores predictions of Gen2 averages of gender standardized AFI are displayed in Table \ref{table_Mean_Joint_Intelligence_Mean_Child_AFI_9}. Again, three models based on Gen2 categories are displayed: Mixed (n $= 337$), Daughters(n $= 260$), and the Sons(n $= 278$). All three models reveal similar results. Gen1 (maternal) intelligence was significantly associated with Gen2 AFI (p $< .01$), while Gen2 (child) intelligence was not significantly associated with Gen2 AFI. A one unit increase in the average standardized intelligence of the children's mothers predicted $\approx .014$ standard deviation increase in average Gen2 AFI. The adjusted R$^{2}$ varied slightly by Gen2 categories (Mixed $= .086$, Daughters $= .097$, Sons $= .100$); each was practically identical to the Mean Gen1 Intelligence models.\pagebreak
\begin{landscape}
\begin{longtable}{@{\extracolsep{5pt}}lccc} 
\caption{Gen1 Mean Intelligence $\rightarrow$ Gen2 Mean AFI}\label{table_Mean_Mom_Intelligence_Mean_Child_AFI_9}
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
& \multicolumn{3}{c}{\textit{Dependent variable:} Average of Gen2 AFI} \\ 
\cline{2-4}
\partialinput{10}{22}{../Common/content/tables/table_Mean_Mom_Intelligence_Mean_Child_AFI_9.tex}
\end{longtable}\pagebreak

>>>>

\subsubsection{Gen1 Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Gen1 sister averages (NLSY79 mothers) of standardized AFQT scores were used to predict Gen2 averages of gender-standardized AFI. Table \ref{table_Mean_Mom_Intelligence_Mean_Child_AFI_9} displays the results by Gen2 categories. The Mixed model reports the averages of the first born child (both males and females) of each sister (n $= 342$), the Daughters model reports the averages of the first born girls (n $= 264$), and the Sons model reports the averages of the first born sons (n $= 282$). All three models show similar results. A one unit increase in the average standardized intelligence of the children's mothers predicted a statistically significant increase of $\approx .013$ standard deviations in average Gen2 AFI. The adjusted R$^{2}$ varied slightly by Gen2 categories (Mixed $= .087$, Daughters $= .097$, Sons $= .103$).

\subsubsection{Gen2 Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Gen2 cousin averages of standardized intellectual ability scores were used to predict Gen2 averages of gender-standardized AFI. Table \ref{table_Mean_Child_Intelligence_Mean_Child_AFI_9} displays the results by Gen2 categories. The Mixed model reports the averages of the first-borns of each of the NLSY79 mothers (sisters) (n $= 344$), the Daughters model reports the averages of the first born girls (n $= 267$), and the Sons model reports the averages of the first born sons (n $= 283$). All three models show similar results. A one unit increase in the average standardized intelligence of the children predicted a statistically significant $\approx .075$ standard deviation increase in average Gen2 AFI. The adjusted R$^{2}$ varied slightly by Gen2 categories (Mixed $= .014$, Daughters $= .016$, Sons $= .009$).

\subsubsection{Joint Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Results from the Gen1 sister averages of standardized AFQT scores and Gen2 cousin averages of standardized intellectual ability scores predicting Gen2 averages of gender-standardized AFI are displayed in Table \ref{table_Mean_Joint_Intelligence_Mean_Child_AFI_9}. Again, three models based on Gen2 categories are displayed: Mixed (n $= 337$), Daughters(n $= 260$), and the Sons(n $= 278$). All three models reveal similar results. Gen1 (maternal) intelligence was significantly associated with Gen2 AFI (p $< .01$), while Gen2 (child) intelligence was not significantly associated with Gen2 AFI. A one unit increase in the average standardized intelligence of the children's mothers predicted $\approx .014$ standard deviation increase in average Gen2 AFI. The adjusted R$^{2}$ varied slightly by Gen2 categories (Mixed $= .086$, Daughters $= .097$, Sons $= .100$); each was practically identical to the Gen1 Mean Intelligence models.

Moreover, when we broaden our sample to all Mother-Child pairs, we see that the relationship between Gen2 AFI and Gen2 general ability is small $(r =.139)$ and smaller than the relationship between Gen2 AFI and Gen1 general ability $(r=.215$; see Figure \ref{plot_gen2_afi}).\pagebreak
\begin{landscape}
\begin{longtable}{@{\extracolsep{5pt}}lccc} 
\caption{Gen1 Mean Intelligence $\rightarrow$ Gen2 Mean AFI}\label{table_Mean_Mom_Intelligence_Mean_Child_AFI_9}
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
& \multicolumn{3}{c}{\textit{Dependent variable:} Average of Gen2 AFI} \\ 
\cline{2-4}
\partialinput{10}{22}{../Common/content/tables/table_Mean_Mom_Intelligence_Mean_Child_AFI_9.tex}
\end{longtable}\pagebreak

### 2016-01-12 Common/content/writing/results_within.tex
fixed tables

We replicated the between-family analyses reported in the previous subsection, using within-family difference scores and means. Using the discordant sibling model, we predicted the differences in Generation 2 AFI as a function of differences in intelligence, controlling for means of the outcomes and predictors. We ran three series of models, where we examined the individual and then joint influence of Gen1 intelligence and Gen2 intelligence. Moreover, within each series we included three Generation 2 linking method variants, just as we did in the between family analyses: Mixed model reports the differences of the first borns of each sister, the Daughters model reports the differences of the first born girls, and the Sons model reports the differences of the first born sons.

@@

>>>>

We replicated the between-family analyses reported in the previous subsection, using within-family difference scores and means. Using the discordant sibling model, we predicted the differences in Generation 2 AFI as a function of differences in intelligence, controlling for means of the outcomes and predictors. We ran three series of models, where we examined the individual and then joint influence of Gen1 intelligence and Gen2 intelligence. Moreover, within each series we included three Generation 2 linking method variants, just as we did in the between family analyses: the Mixed model reports the differences of the first borns of each sister, the Daughters model reports the differences of the first born girls, and the Sons model reports the differences of the first born sons.

@@

Moreover, when we replicate this analysis using the larger Gen1 sample of siblings, we find that the between-family relationship of Gen1 AFI and Gen1 general ability is moderate $(r =.291)$, but that within the family, the smarter sibling (as illustrated by the blue marginal distributions in Figure \ref{plot_gen1_afi}) were indistinguishable from the less smart sibling (the red marginal distributions). Had there been a within-family effect of intelligence on AFI, the marginal distributions of AFI would not overlap.\pagebreak
